Performance benefits of Intel\textsuperscript{\textregistered}~Optane\texttrademark~DC persistent memory for the parallel processing of large neuroimaging data

### 2019-12-17 paper.tex
requested changes

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
\algnewcommand\algorithmicforeach{\textbf{for each}}
\algdef{S}[FOR]{ForEach}[1]{\algorithmicforeach\ #1\ \algorithmicdo}
\algblock{Input}{EndInput}
\algnotext{EndInput}
\algblock{Output}{EndOutput}
\algnotext{EndOutput}
\newcommand{\Desc}[2]{\State \makebox[2em][l]{#1}#2}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{xspace}
\usepackage{multirow}
\usepackage[binary-units]{siunitx}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\newcommand{\bigdata}{Big Data\xspace}
\newcommand{\bigbrain}{BigBrain\xspace}

@@

\title{Performance benefits of Intel Optane DC storage for the parallel processing of large neuroimaging data}

@@

\begin{abstract}
    Open-access neuroimaging datasets have reached up to $\SI{}{\peta\byte}$ of data, and
    continue to grow. The ability to leverage the entirety of these datasets are
    limited to a restricted number of labs with both the capacity and infrastructure
    to process the data. Whereas BigData engines have significantly reduced
    application performance penalties with respect to data movement, their applied
    strategies are not necessarily practical within neuroimaging workflows where
    intermediary results may need to be materialized to shared storage for post-processing
    analysis. In this paper we evaluate the performance advantage brought by Intel
    Optane DC persistent memory storage for the processing of large neuroimaging 
    datasets using the two available configurations modes: Memory Mode and App Direct Mode.
    We employ a synthetic algorithm on the $\SI{76}{\gibi\byte}$ \bigbrain
    and a real neuroimaging application on the Consortium for Reliability and Reproducibility (CoRR)
    dataset using 25 and 96 parallel
    processes in both cases. Our results show that performance of applications executing on
    Optane is superior to that of other storage devices, with the exception of DRAM. This is
    the case in both Memory and App Direct Mode and irrespective of the amount of data and 
    parallelism. Furthermore, Optane in App Direct Mode is believed to benefit from the use
    of memory as a cache for writing in cases where output data is significantly smaller than
    available memory. We believe the use of Optane will be beneficial to both neuroimaging applications
    running on HPC or visualization of large, high-resolution images.
\end{abstract}

@@

Due to storage limitations, only subsets of such neuroimaging repositories 
can be processed in a typical research laboratory. Moreover, as these datasets 
are extremely large and are only increasing in size, they are typically stored 
in higher-capacity, slower storage devices, such as hard disk drives, or external
parallel file systems, such as Lustre. In such conditions, large-scale 
studies in neuroimaging remain limited to labs with access to adequate 
infrastructure.

@@

To mitigate the effects of data writes, kernel-based
strategies,such as the writeback cache, have been developed. The writeback cache
allows processes to use the memory as a cache for writing. The cache size,
however,
is configurable, but nevertheless limited. When writes approach the cache's capacity,
processes performing writes start to be throttled. After capacity is reached,
processes can no longer use the cache until all written data on cache is flushed
to the appropriate storage device.

Whereas Operating Systems and popular Big Data engines, such as MapReduce~\cite{mapreduce} and Apache Spark~\cite{spark}, have
incorporated software solutions to limit data transfers (e.g. writeback cache, in-memory computing,
data locality, and lazy evaluation), hardware has also adapted to the growing datasets.
One such improvement is the concept of placing flash storage directly on the Dual
In-line Memory Module (DIMM), thereby reducing the latency of accessing data on 
flash storage devices. While the latency of these devices is improved, the bandwidth
remains the same. However, as noted by \cite{nvdimms}, a severe performance degradation
can be experienced by having memory traffic and I/O placed on the same bus.

Intel Optane DC Persistent Memory storage~\cite{optanebrief} is a recently introduced high-performance
storage technology that also sits on the DIMMs to reduce latency to the device.
Unlike DRAM DIMMs, it can accommodate twice the volume ($\SI{512}{\giga\byte}$) of
storage enabling it to be a more cost effective alternative to DRAM storage.
There are two principle configuration modes, namely Memory Mode and App Direct Mode,
that enable the storage to either be accessed as an extension of available memory 
or as a non-volatile memory storage device.

In this paper, we aim to:
\begin{itemize}
        \item Quantify the added value of Intel Optane DC Persistent Memory on 
            processing large neuroimaging data using representative pipelines; and
        \item Determine when a given Intel Optane storage configuration (Memory 
            and App Direct Mode) is preferable.
\end{itemize}

@@

The server used for the experiments consisted of 12 \SI{64}{\gibi\byte} DRAM devices,
resulting in a total of \SI{768}{\gibi\byte} of DRAM, and 12 \SI{256}{\giga\byte} Intel Optane DC Memory
devices, resulting in total of \SI{3}{\tebi\byte} of Intel Optane DC Memory. The theoretical maximum bandwidth
of DRAM is estimated to be around \SI{275}{\gibi\byte}~/s whereas for Optane, it is
estimated to be around \SI{250}{\gibi\byte}~/s.
Other storage devices included a \SI{240}{\gibi\byte} Micron SATA SSD, of which
only \SI{149}{\gibi\byte} where available for the experiments, and a
\SI{720}{\tebi\byte} shared Dell EMC Isilon cluster. The Isilon cluster is an nsf4 mounted
cluster (10Gb network) consisting of 5 nodes of 36 hard disk drives (HDD). The local disk
was the only storage device set up as a writeback device. Both Isilon and Optane in
App Direct mode were configured as write-through devices, meaning they did not leverage
DRAM as a cache for writes. This was achieved in Optane by configuring the xfs filesystem with
DAX. All storage devices were benchmarked using the following \href{https://github.com/ValHayot/paper-memory-storage/scripts/bench_disks.sh}{script}.
The result of the benchmarks can be seen in Table \ref{table:bandwidths}.

@@

\begin{table}
\begin{center}
 \begin{tabular}{ |c|c|c| } 
     \cline{2-3}
     \multicolumn{1}{c|}{} & \multicolumn{2}{c|}{Bandwidth (MB/s)} \\\hline
  Device & Read & Write \\
 \hline
 DRAM & 5304.3 & 3338.2 \\  
 Optane & 3379.2 & 2396.2 \\   
 Local Disk & 518.6 & 240.4 \\
 Isilon & 117.0 & 111.8 \\
 \hline
\end{tabular}\caption{Perceived read and write bandwidth of each storage device}\label{table:bandwidths}
\end{center}
\end{table}

@@

Memory Mode leverages Intel Optane DC persistent memory (Optane) to extend the system's available
memory. In this mode, Optane uses DRAM as a cache and is accessible as
volatile addressable main memory. By extending main memory, Memory Mode enables the
fast access of large volumes of data.

While enabling the access to larger amount of memory than typically accessible otherwise,
it is anticipated that Memory Mode will be slower than AD mode for all memory accessed that
are not cached in DRAM. We evaluated all available devices (Optane through tmpfs, local disk and Isilon)
in Memory Mode.

@@

App Direct Mode enables Optane to be accessed as a high-performance storage device.
Unlike Memory Mode, the OS is able to differentiate between DRAM and Optane,
treating them as two distinct entities. As Optane is located on
the DIMMs, it exhibits lower latencies than traditional SSDs, while being able
to be accessed like an SSD. Optane does not use the DRAM as cache in App Direct Mode
if configured with DAX.

Similarly to Memory Mode, our experiments evaluated all available filesystems in App Direct
Mode. In other words, we evaluated DRAM performance through tmpfs, Optane performance through
a \SI{1.5}{\tebi\byte} Intel Optane DC Persistent Memory storage device and local disk
and Isilon using the same storage mounts available in Memory Mode.

@@

For applications which have negligible CPU time, as is the case with our experiments, 
it is expected that the I/O duration can estimate the total makespan. However, there
are certain instances where the makespan may be below the I/O duration. This is expected
to occur with storage devices that scale well with increased parallelism, as should be the
case for DRAM, Optane and Isilon, which all consist of multiple devices. Should the
scalable storage devices predict the makespan accurately, it is believed that other overheads
would be at play making the device unfavourable for parallel processing.

@@

\begin{algorithm}\caption{Incrementation}\label{alg:incrementation}
    \begin{algorithmic}[1]
    \Input
        \Desc{$x$}{a sleep delay in seconds}
        \Desc{$n$}{a number of iterations}
        \Desc{$C$}{a set of image chunks}
        \Desc{$fs$}{filesystem to write to (tmpfs, Optane, local disk, Isilon)}
    \EndInput
    \ForEach{$chunk \in C$}
        \State read $chunk$ from $fs$
        \State $chunk\gets chunk+1$
        \State save $chunk$ to $fs$
    \EndFor
    \end{algorithmic}
\end{algorithm}

Due to the size and uniqueness of the \bigbrain, standardized processing pipelines
have yet to be developed. In order to quantify the effects of storage devices
on such a dataset, we have implemented a naive synthetic application that takes 
the image, split into blocks, and increments all the voxels within each block, in parallel (Algorithm~\ref{alg:incrementation}).
This application
enables us to read and write to different storage devices and ensure that written
data could not have been previously cache
in-memory, by ensuring that read and written data are not the same. This application
was parallelized us in two different ways: using both GNU Parallel and Apache Spark 2.4.3 (PySpark). 
All code was implemented in Python 3.6.

There are some notable differences between the GNU Parallel and Apache Spark implementations.
For instance, in the GNU Parallel implementation, all operations (i.e. reading, incrementing and writing)
are done within the same task. The same task is applied to all the \bigbrain blocks and GNU Parallel is 
provided with all the different command-line calls, processing a subset of them at a time,
depending on the level of parallelism provided. Reading in the GNU Parallel implementation 
is achieved using the popular neuroimaging I/O: Nibabel. When provided with a filename, as is the case
for the GNU Parallel implementation, Nibabel will simply load the header in memory, and memmap the data
using Numpy. This step will be referred to as ``load header''. It is only when the data is actually required
(i.e. during incrementation), that the data will be loaded into memory.

The Apache Spark implementation differs from that of GNU Parallel in that scheduling decisions
were entirely differed to Spark. That is, rather than simply parallelizing the filenames, to be
loaded in a similar fashion to that of GNU Parallel, we chose to load the using Spark's built-in
\texttt{BinaryFiles}, which loads whole files, in binary format, into Spark partitions. As the
data would have been pre-loaded by Spark, ``loading header'' only measures the time to
convert the binary images into Nibabel objects, and ``incrementation'' consists purely
of the time it took to increment the data. Therefore, only the write time would be
comparable between the two implementations, as it is achieved using the same method.
Furthermore, the Spark implementation also differs in that read, increment and write
were separated into three map tasks, which can enable shuffling between the tasks should 
it be determined as necessary by the Spark scheduler.

@@

For the 40~$\mu$m \bigbrain, we have executed the application using GNU Parallel for parallelization
using 25 and 96 processes on the 125 40$\mu$m \bigbrain blocks. Data was read and written to either
DRAM (App Direct only), Optane, local disk and Isilon.

For the 20~$\mu$m \bigbrain, we used the same application, however in this case, it was executed
using both Apache Spark and GNU Parallel. The configuration was also generally the same, having 
experiments using both 25 and 96 cpus. Repeating the experiments using a larger dataset would enable
us to determine the effects when Optane would have to be partially relied upon due to insufficient DRAM
space. Moreover, with such large datasets, it is likely that Big Data frameworks would be used rather than
more traditional parallelism frameworks. Using Spark, we can evaluate how Big Data Frameworks perform on
different storage devices.

Since storage was limited on DRAM and local disk and was not large enough to process the entire
20~$\mu$m \bigbrain, these devices were omitted in the processed of this dataset.

@@

The BIDS App Example is a template example for creating a Brain Imaging Data Structure (BIDS)
compliant application. It runs a standard neuroimaging brain extraction application (FSL BET) 
on all the anatomical (T1W) images of dataset containing numerous subjects. This step is 
referred to as Participant analysis within the BIDS App Example. An optional
step of the BIDS App example, referred to as Group analysis, computes the average brain
mask size of the entire dataset.

For our experiments, we used the entire CoRR dataset, made available on \href{}{DataLad} and
applied Participant analysis to it. The BIDS App Example was executed using a Singularity container stored on
Isilon. As in the BigBrain Incrementation, the experiment parallelized 
using GNU Parallel with 25 and 96 processes. The conditions were executed in both App Direct and Memory Mode
using Isilon, local disk and Optane as storage. Each experiment was repeated 3x.

@@

\subsection{40~$\mu$m \bigbrain incrementation}
\subsubsection{25 Processes}
\begin{figure}
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/makespan-real-40bb_25cpus.pdf}
    \captionsetup{belowskip=-10pt}
    \caption{Makespan of the \bigbrain Incrementation application using 25 
             processes on the 125 blocks on the 40~$\mu$m image given different
             storage configurations. Anticipated read and write duration was 
             measured by applying the perceived bandwidths Table~\ref{table:bandwidths}
             to Eq.~\ref{eq:makespan}. 3 
             repetitions were performed. *Local SSD did not complete the
    writing of the last few blocks due to storage limitations.}\label{fig:makespan-25cpus}
\end{figure}
\begin{figure}
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/stacked-real-40bb_25cpus.pdf}
    \captionsetup{belowskip=0pt}
    \caption{Total read/increment/write breakdown of \bigbrain Incrementation using 25 cpus. 3 repetitions were performed. *Local SSD did not
             complete the writing of the last few blocks due to storage limitations}\label{fig:stacked-25cpus}
\end{figure}
\begin{figure*}
    \begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/gantt-1572380543-tmpfsAD_1it_25cpus_40bb-1.pdf}
    \caption{DRAM}
\end{subfigure}
\begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/gantt-1572382482-optaneAD_1it_25cpus_40bb-1.pdf}
    \caption{Optane}
\end{subfigure}
\begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/gantt-1572382794-localAD_1it_25cpus_40bb-1.pdf}
    \caption{local SSD*}
\end{subfigure}
\begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/gantt-1572388118-isilonAD_1it_25cpus_40bb-2.pdf}
    \caption{Isilon}\label{fig:gantt25isilon}
\end{subfigure}
    \captionsetup{belowskip=-10pt}
\caption{Gantt charts for each storage device (App Direct Mode) processing 125 blocks of the 40$\mu$m BigBrain using 25 processes. *Some local SSD writes did not complete due to storage limitations.}\label{fig:gantt25}
\end{figure*}

As can be seen in Figure~\ref{fig:makespan-25cpus}, DRAM (App Direct Mode) is the 
most efficient, with a makespan of 7.6s. Optane in Memory mode is a close second
with a makespan of approximately 10s. Optane in App Direct mode is approximately 
12x slower than Optane in Memory Mode. Isilon in both Memory Mode and App Direct 
Mode is more than 120x slower than Optane in Memory Mode. Other than for Optane, 
there is no significant difference between Memory Mode and App Direct executions.

When comparing the makespan to the estimated read and write duration, local SSD
in both modes and Optane in App Direct Mode have longer makespans than the anticipated.
Optane was approximately 2.2x slower than expected, whereas local SSD was approximately
1.4x slower than expected, in both modes. For those where the I/O time was overestimated,
DRAM was almost 5x faster in reality compared to what was estimated. Optane in Memory Mode
had a smaller difference, with it being 3.8x faster than expected. Isilon in App Direct
mode had a larger difference than Memory mode, with it being 1.1x faster than expected.

The total task duration breakdowns for each device (Figure~\ref{fig:stacked-25cpus})
shows that, as expected, I/O times vary greatly between devices, with DRAM having the
best total read and write speeds. Optane in Memory Mode is very close to DRAM in speed,
whereas Optane in App Direct mode 7x slower in terms of reading and 23x slower in terms
of writing. Local SSD, on the other hand, exhibits similar write times as DRAM, but is
approximately 96x that DRAM with respect to reads. Header loading for local SSD is also 90x longer
than that of DRAM, however, remains negligible compared to other tasks. Isilon, on the 
other hand, has non negligible header load times, with it being more than 5000 times
slower than that of DRAM. Interestingly enough, the read/increment times on Isilon are
slower than that of the local SSD, although Isilon makes up for its speedier reads with 
writes that are around 184x slower than DRAM.

The App Direct mode Gantt charts (Figure~\ref{fig:gantt25}) also reflect what is observed in Figure~\ref{fig:stacked-25cpus}.
DRAM is significantly faster than all other storage, leading it to barely appear within the Gantt chart.
DRAM was measured to have an average of 17.2 parallel tasks throughout the execution. Optane takes
significantly longer than DRAM and appears to spend a significant amount of time writing. The average
parallelism measured for Optane was 24. Unlike Optane, local SDD spends the vast majority of its time
reading, while spending a negligible amount of time writing. Like Optane, it averages 24 parallel tasks
throughout its execution. Unlike the other three storages, Isilon shows spacing between the loading of the header,
read and increment and write. Isilon was measured to have an average parallelism of 14. Overall, each storage device
execution displayed that some tasks took longer than others.

>>>>

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
\algnewcommand\algorithmicforeach{\textbf{for each}}
\algdef{S}[FOR]{ForEach}[1]{\algorithmicforeach\ #1\ \algorithmicdo}
\algblock{Input}{EndInput}
\algnotext{EndInput}
\algblock{Output}{EndOutput}
\algnotext{EndOutput}
\newcommand{\Desc}[2]{\State \makebox[2em][l]{#1}#2}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{xspace}
\usepackage{multirow}
\usepackage{textcomp}
\usepackage[binary-units]{siunitx}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\newcommand{\bigdata}{Big Data\xspace}
\newcommand{\bigbrain}{BigBrain\xspace}

@@

\title{Performance benefits of Intel\textsuperscript{\textregistered}~Optane\texttrademark~DC persistent memory for the parallel processing of large neuroimaging data}

@@

\begin{abstract}
    Open-access neuroimaging datasets have reached petabyte scale, and continue to grow.
The ability to leverage the entirety of these datasets is limited to a restricted
number of labs with both the capacity and infrastructure to
process the data. Whereas Big Data engines have significantly
reduced application performance penalties with respect to data
movement, their applied strategies are not necessarily practical
within neuroimaging workflows where intermediary results may
need to be materialized to shared storage for post-processing
analysis. In this paper we evaluate the performance advantage
brought by Intel\textsuperscript{textregistered}~Optane\texttrademark~DC persistent memory for the
processing of large neuroimaging datasets using the two available
configurations modes: Memory Mode and App Direct Mode. We
employ a synthetic algorithm on the 76 GiB and 603 GiB BigBrain, as well as apply a
standard neuroimaging application on the Consortium for Reliability
and Reproducibility (CoRR) dataset using 25 and 96 parallel
processes in both cases. Our results show that the performance of
the applications leveraging persistent memory is superior to that of other
storage devices, with the exception of DRAM. This is the case
in both Memory and App Direct Mode and irrespective of the
amount of data and parallelism. Furthermore, persistent memory in App
Direct Mode is believed to benefit from the use of memory as
a cache for writing in cases where output data is significantly
smaller than available memory. We believe the use of persistent memory
will be beneficial to both neuroimaging applications running on
HPC or visualization of large, high-resolution images.
\end{abstract}

@@

Due to storage limitations, only subsets of such neuroimaging repositories 
can be processed in a typical research laboratory. Moreover, as these datasets 
are extremely large and are only increasing in size, they are typically stored 
in higher-capacity, slower storage devices, such as hard disk drives, or external
parallel file systems, such as Lustre. In such conditions, large-scale 
studies in neuroimaging remain limited to labs with access to adequate 
infrastructures.

@@

To mitigate the effects of data writes, kernel-based
strategies, such as the writeback cache, have been developed. The writeback cache
allows processes to use the memory as a cache for writing. The cache size,
however,
is configurable, but nevertheless limited. When writes approach the cache's capacity,
processes performing writes start to be throttled. Once the cache capacity is reached,
processes can no longer use the cache until all cached written data is flushed
to the appropriate storage device.

Whereas Operating Systems and popular Big Data engines, such as MapReduce~\cite{mapreduce} and Apache Spark~\cite{spark}, have
incorporated software solutions to limit data transfers (e.g. writeback cache, in-memory computing,
data locality, and lazy evaluation), hardware has also adapted to the growing datasets.
One such improvement is the concept of placing persistent storage directly on the Dual
In-line Memory Module (DIMM), thereby reducing the latency of accessing data on 
flash storage devices. While the latency of these devices is improved, the bandwidth
remains the same. However, as noted by \cite{nvdimms}, a severe performance degradation
can be experienced by having memory traffic and I/O placed on the same bus.

Intel Optane DC Persistent Memory~\cite{optanebrief} is a high-performance
storage technology that resides on the DIMMs to reduce latency to the device.
The first generation of Intel Optane DC persistent memory offers capabities of 128~GB,
256~GB, and 512~GB, enabling it to be more cost effective than DRAM storage.
There are two configuration modes, Memory Mode and App Direct Mode,
that enable the storage to either be accessed as an extension of volatile main memory
or as a non-volatile memory storage device\footnote{https://itpeernetwork.intel.com/intel-optane-dc-persistent-memory-operating-modes/}.

In this paper, we aim to:
\begin{itemize}
        \item Quantify the added value of Intel Optane DC persistent memory on 
            processing large neuroimaging data using representative pipelines; and
        \item Determine when a given Intel Optane DC persistent memory configuration (Memory 
            and App Direct Mode) is preferable.
\end{itemize}

@@

The server used for the experiments consisted of 12X~\SI{64}{\gibi\byte} DRAM devices,
resulting in a total of \SI{768}{\gibi\byte} of DRAM, and 12X~\SI{256}{\giga\byte} Intel Optane DC persistent memory
modules, resulting in total of \SI{3}{\tebi\byte} of Intel Optane DC persistent memory.
Other storage devices included a \SI{240}{\gibi\byte} Micron SATA SSD, of which
only \SI{149}{\gibi\byte} where available for the experiments, and a
\SI{720}{\tebi\byte} shared Dell EMC Isilon cluster. The Isilon cluster is an nsf4 mounted
cluster (10Gb network) consisting of 5 nodes of 36 hard disk drives (HDD). The local disk
was the only storage device set up as a writeback device. Both Isilon and Optane DC persistent memory in
App Direct mode were configured as write-through devices, meaning they did not leverage
DRAM as a cache for writes. For Optane DC persistent memory, this was achieved by configuring the xfs filesystem with
DAX. All storage devices were benchmarked using the following \href{https://github.com/ValHayot/paper-memory-storage/scripts/bench_disks.sh}{script}.
The result of the benchmarks can be seen in Table \ref{table:bandwidths}.

@@

\begin{table}
\begin{center}
 \begin{tabular}{ |c|c|c| } 
     \cline{2-3}
     \multicolumn{1}{c|}{} & \multicolumn{2}{c|}{Bandwidth (MB/s)} \\\hline
  Device & Read & Write \\
 \hline
 DRAM & 5304.3 & 3338.2 \\  
 Optane & 3379.2 & 2396.2 \\   
 Local Disk & 518.6 & 240.4 \\
 Isilon & 117.0 & 111.8 \\
 \hline
\end{tabular}\caption{Measured read and write bandwidth of each storage device}\label{table:bandwidths}
\end{center}
\end{table}

@@

Memory mode leverages Intel Optane DC persistent memory (DCPMM) to extend the system's available
memory. In this mode, Optane DCPMM uses DRAM as a cache and is accessible as
volatile addressable main memory. By extending main memory, Memory mode enables the
fast access of large volumes of data. For instance, Memory mode on our server enabled use of
3~TiB of main memory, whereas it would have only had 768~GiB of main memory if Optane DCPMM could not be
leveraged. Although DRAM is used as cache by Optane DCPMM, it is not visible to the operating system.

While enabling access to a larger amount of memory than typically accessible otherwise,
it is anticipated that Memory mode will be slower than App Direct mode for all memory accessed that
are not cached in DRAM. We evaluated all available devices (Optane DCPMM through tmpfs, local disk and Isilon)
in Memory mode.

@@

App Direct mode enables Optane DCPMM to be accessed as a high-performance storage device.
Unlike Memory mode, the OS is able to differentiate between DRAM and Optane DCPMM,
treating them as two distinct memory tiers. Optane DCPMM
does not use the DRAM as cache in App Direct Mode if configured with DAX.

Similarly to Memory Mode, our experiments evaluated all available filesystems in App Direct
Mode. In other words, we evaluated DRAM (tmpfs mount), Optane DCPMM (\SI{1.5}{\tebi\byte} persistent
memory mount), local SSD and Isilon.

@@

For applications which have negligible CPU time, as is the case with our experiments, 
it is expected that the I/O duration can estimate the total makespan. However, there
are certain instances where the makespan may be below the I/O duration. This is expected
to occur with storage devices that scale well with increased parallelism, as should be the
case for DRAM, Optane DCPMM and Isilon, which all consist of multiple devices. Should the
scalable storage devices predict the makespan accurately, it is believed that other overheads
would be at play making the device unfavourable for parallel processing.

@@

\begin{algorithm}\caption{Incrementation}\label{alg:incrementation}
    \begin{algorithmic}[1]
    \Input
        \Desc{$x$}{a sleep delay in seconds}
        \Desc{$n$}{a number of iterations}
        \Desc{$C$}{a set of image chunks}
        \Desc{$fs$}{filesystem to write to (tmpfs, Optane DCPMM, local disk, Isilon)}
    \EndInput
    \ForEach{$chunk \in C$}
        \State read $chunk$ from $fs$
        \State $chunk\gets chunk+1$
        \State save $chunk$ to $fs$
    \EndFor
    \end{algorithmic}
\end{algorithm}

Due to the size and uniqueness of the \bigbrain, standardized processing pipelines
have yet to be developed. In order to quantify the effects of storage devices
on such a dataset, we have implemented a naive synthetic application that takes 
the image, split into blocks, and increments all the voxels within each block, in parallel (Algorithm~\ref{alg:incrementation}).
This application
enables us to read and write to different storage devices and ensure that written
data could not have been previously cache
in-memory, by ensuring that read and written data are not the same. This application
was parallelized in two different ways: using both GNU Parallel and Apache Spark 2.4.3 (PySpark). 
All code was implemented in Python 3.6.

There are some notable differences between the GNU Parallel and Apache Spark implementations.
For instance, in the GNU Parallel implementation, all operations (i.e. reading, incrementing and writing)
are done within a single task. This task is applied to all the \bigbrain blocks using GNU Parallel,
processing a subset of them at a time,
depending on the level of parallelism provided. Reading in the GNU Parallel implementation 
is achieved using the popular neuroimaging I/O: \href{https://nipy.org/nibabel/}{NiBabel}. When provided with a filename, as is the case
for the GNU Parallel implementation, Nibabel will simply load the header in memory, and memory map the data
using \href{https://numpy.org/}{NumPy}. This step will be referred to as ``load header''. It is only when the data is actually required
(i.e. during incrementation), that the data will be loaded into memory.

The Apache Spark implementation differs from that of GNU Parallel in that scheduling decisions
were entirely differed to Spark. That is, rather than simply parallelizing over the \bigbrain block filenames,
we opted to read the data using Spark's built-in
\texttt{BinaryFiles}, loading whole files, in binary format, into Spark partitions. As the
data would have been pre-loaded by Spark, ``loading header'' only measures the time to
convert the binary images into NiBabel objects, and ``incrementation'' consists solely
of the time it took to increment the data. Therefore, only the write time would be
comparable between the two implementations, as it is achieved using the same method.
Furthermore, the Spark implementation also differs in that read, increment and write
were separated into three map tasks, which can enable shuffling between the tasks should 
it be determined as necessary by the Spark scheduler.

@@

For the 40~$\mu$m \bigbrain, we have executed the application using GNU Parallel for parallelization
using 25 and 96 processes on the 125 40$\mu$m \bigbrain blocks. Data was read and written to either
DRAM (App Direct only), Optane DCPMM, local disk and Isilon.

For the 20~$\mu$m \bigbrain, we used the same application, however in this case, it was executed
using both Apache Spark and GNU Parallel. The configuration was also generally the same, having 
experiments using both 25 and 96 cpus. Repeating the experiments using a larger dataset would enable
us to determine the effects when Optane DCPMM would have to be partially relied upon due to insufficient DRAM
space. Moreover, with such large datasets, it is likely that Big Data frameworks would be used rather than
more traditional parallelism frameworks. Using Spark, we can evaluate how Big Data Frameworks perform on
different storage devices.

Since storage was limited on DRAM and local disk and was not large enough to process the entire
20~$\mu$m \bigbrain, these devices were omitted in the processing of this dataset.

@@

The BIDS App Example is a template example for creating a Brain Imaging Data Structure (BIDS)
compliant application. It runs a standard neuroimaging brain extraction application (FSL BET) 
on all the anatomical (T1W) images of dataset containing numerous subjects. This step is 
referred to as Participant analysis within the application. An optional
step of the BIDS App example, referred to as Group analysis, computes the average brain
mask size of the entire dataset.

For our experiments, we used the entire CoRR dataset, made available on \href{}{DataLad} and
applied Participant analysis to it. The BIDS App Example was executed using a Singularity container stored on
Isilon. As in the BigBrain Incrementation, the experiment was parallelized 
using GNU Parallel with 25 and 96 processes. The conditions were executed in both App Direct and Memory Mode,
where Isilon and the local SSD were only evaluated in Memory Mode. Each experiment was repeated 3x.

@@

\subsection{40~$\mu$m \bigbrain incrementation}
\subsubsection{25 Processes}
\begin{figure}
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/makespan-real-40bb_25cpus.pdf}
    \captionsetup{belowskip=-10pt}
    \caption{Makespan of the \bigbrain Incrementation application using 25 
             processes on the 125 blocks on the 40~$\mu$m image given different
             storage configurations. Anticipated read and write duration was 
             measured by applying the perceived bandwidths Table~\ref{table:bandwidths}
             to Eq.~\ref{eq:makespan}. 3 
             repetitions were performed. *Local SSD did not complete the
    writing of the last few blocks due to storage limitations.}\label{fig:makespan-25cpus}
\end{figure}
\begin{figure}
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/stacked-real-40bb_25cpus.pdf}
    \captionsetup{belowskip=0pt}
    \caption{Total read/increment/write breakdown of \bigbrain Incrementation using 25 cpus. 3 repetitions were performed. *Local SSD did not
             complete the writing of the last few blocks due to storage limitations}\label{fig:stacked-25cpus}
\end{figure}
\begin{figure*}
    \begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/gantt-1572380543-tmpfsAD_1it_25cpus_40bb-1.pdf}
    \caption{DRAM}
\end{subfigure}
\begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/gantt-1572382482-optaneAD_1it_25cpus_40bb-1.pdf}
    \caption{Optane DCPMM}
\end{subfigure}
\begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/gantt-1572382794-localAD_1it_25cpus_40bb-1.pdf}
    \caption{local SSD*}
\end{subfigure}
\begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{./experiments/bigbrain-incrementation/figures/gantt-1572388118-isilonAD_1it_25cpus_40bb-2.pdf}
    \caption{Isilon}\label{fig:gantt25isilon}
\end{subfigure}
    \captionsetup{belowskip=-10pt}
\caption{Gantt charts for each storage device (App Direct Mode) processing 125 blocks of the 40$\mu$m BigBrain using 25 processes. *Some local SSD writes did not complete due to storage limitations.}\label{fig:gantt25}
\end{figure*}

As can be seen in Figure~\ref{fig:makespan-25cpus}, DRAM (App Direct mode) is the 
most efficient, with a makespan of 7.6s. Optane in Memory mode is a close second
with a makespan of approximately 10s. As caching was disabled in App Direct mode, 
Optane DCPMM in this mode is approximately 
12x slower than Optane in Memory Mode. Isilon in both Memory mode and App Direct 
mode is more than 120x slower than Optane DCPMM in Memory mode. Other than for Optane DCPMM, 
there is no significant difference between Memory mode and App Direct executions.

When comparing the makespan to the estimated read and write duration, local SSD
in both modes and Optane DCPMM in App Direct mode have longer makespans than anticipated.
Optane DCPMM was approximately 2.2x slower than expected, whereas local SSD was approximately
1.4x slower than expected, in both modes. For those where the I/O time was overestimated,
DRAM was almost 5x faster in reality compared to what was estimated. Optane DCPMM in Memory Mode
had a larger difference, with it being 3.8x faster than expected. Isilon in App Direct
mode had a smaller difference than Memory mode, with it being 1.1x faster than expected.

The total task duration breakdowns for each device (Figure~\ref{fig:stacked-25cpus})
shows that, as expected, I/O times vary greatly between devices, with DRAM having the
best total read and write speeds. Optane DCPMM in Memory Mode is very close to DRAM in speed,
whereas Optane DCPMM in App Direct mode was 7x slower in terms of reading and 23x slower in terms
of writing. Local SSD, on the other hand, exhibits similar write times as DRAM, but is
approximately 96x slower than DRAM with respect to reads. Header loading for local SSD is also 90x longer
than that of DRAM, however, it remains negligible compared to other tasks. Isilon, on the 
other hand, has non negligible header load times, with it being more than 5000x
slower than that of DRAM. Interestingly enough, the read/increment times on Isilon are
faster than that of the local SSD, although Isilon makes up for its speedier reads with 
writes that are around 184x slower than DRAM.

The App Direct mode Gantt charts (Figure~\ref{fig:gantt25}) also reflect what is observed in Figure~\ref{fig:stacked-25cpus}.
DRAM is significantly faster than all other storage, leading it to barely appear within the Gantt chart.
DRAM was measured to have an average of 17.2 parallel tasks throughout the execution. Optane DCPMM takes
significantly longer than DRAM and appears to spend a significant amount of time writing. The average
parallelism measured for Optane DCPMM was 24. Unlike Optane DCPMM, local SSD spends the vast majority of its time
reading, while spending a negligible amount of time writing. Like Optane DCPMM, it averages 24 parallel tasks
throughout its execution. Unlike the other three storage, Isilon shows spacing between the loading of the header,
read and increment and write. Isilon was measured to have an average parallelism of 14. Overall, each storage device
execution displayed that some tasks took longer than others.
